\newpage
\thispagestyle{plain} 

\begin{center}
    \Large \textbf{ACKNOWLEDGMENTS}
\end{center}
\vspace{0.8cm}

\normalsize
\noindent In the name of Allah, the Most Gracious, the Most Merciful, to whom we attribute the strength and success to complete this research.\par\vspace{0.4cm}

We are profoundly grateful to our supervisor, \textbf{Dr. Nada Mostafa Abdelaziem Mostafa}, for her unwavering patience, invaluable expertise, and continuous guidance. Her insights not only shaped the technical direction of this thesis but have also deeply enriched our broader understanding of computer engineering and intelligent systems.\par\vspace{0.4cm}

Equally, we owe our deepest appreciation to \textbf{Dr. Reham Taher Abdelmajeed Salem Al Magraby}. Her transformative support, constant motivation, and words of wisdom were the driving force behind this project. Without her dedicated mentorship and encouragement, the realization of this work would not have been possible. Truly, she has taught us a great deal, and for that, we are immensely grateful.\par\vspace{0.4cm}

We consider ourselves truly fortunate to have worked under the supervision of such exceptional educators, and we hope this thesis serves as a worthy testament to their efforts and guidance.\par\vspace{0.4cm}

Furthermore, we would like to extend our collective and heartfelt thanks to our parents and families for never failing to encourage us to be the best versions of ourselves. Their endless sacrifices and their efforts in instilling the value of hard work and the importance of knowledge within us have been our greatest foundation.\par\vspace{0.4cm}

Finally, a special note of gratitude to all our siblings and loved ones. Without your constant support and encouragement, we would have never been able to accomplish our goals and bring this project to life.